Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-01-08 supplement.tex
bootstrap section in supplement

@@

>>>>

\subsection{Evaluation of estimated confidence intervals}
\label{sec:confidence-intervals}

We estimated 95\% confidence intervals (CIs) as reported in
Tables~\ref{tab:supp-single-origin}--\ref{tab:supp-merger-with-stem-migration}
by refitting each model to block-bootstrap replicated datasets. While our
LD-based inference approach is expected to be unbiased
\citep{Ragsdale2019-nt,Ragsdale2020-nz}, we tested whether our
bootstrap-estimated CIs contain the true parameters using simulated data from
our two highlighted models (the continuous migration model and the merger with
stem migration model).

To do this, we simulated 500 1Mb regions using the same sample sizes as our
analyzed data, with per-base mutation and recombination rates $u=r=10^{-8}$,
using \texttt{msprime} \citep{Baumdicker2022-mj}. From these simulated regions,
we replicated our bootstrapping procedure by resampling (with replacement) 500
regions. We repeated this 500 times (to obtain 500 bootstrap replicate
datasets). We fit the full (non-bootstrapped) simulated dataset and each of the
500 bootstrap replicates, and compared the inferred parameter values to the
true simulated values (Figures~\ref{fig:supp-sim-BS-continuous-migration} and
\ref{fig:supp-sim-BS-merger-with-stem-migration}).

The distribution of refit values for vast majority of parameters overlapped
with the true simulated value. This included every parameter from the
continuous migration model, and all of the parameters in the merger with stem
migration model except for the size, migration rate, and split time of Stem 1
(which experienced a severe bottleneck in the deep past). Because the true
simulated parameter for the bottleneck size in Stem 1 was at the lower bound
($N=100$), this likely is due to the bottleneck re-inferences failing to
fully converge. Overall, bootstrap-estimated CIs typically contain the true
parameter, although CIs for parameters that lie at an optimization bound
should be treated with caution.

Finally, many features of recent human history were consistently parameterized
across each of our tested models (such as recent migration rates and population
sizes). We evaluated the consistency of these recent parameters across our
different models (Figure~\ref{fig:supp-bootstrap-overlap}). In the case that
the model provides a very poor fit to the data (the single origin model) the
CIs of some parameters do not overlap with those of the other models (e.g.,
Mende population sizes, and some migration rates). Many parameters are largely
consistent across each of the models, showing that inferences of recent history
are largely robust to the details of population structure in the distant past.

@@

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-simulated-continuous_migration-bootstrap-fits.pdf}
    \caption{
        \textbf{Evaluation of confidence intervals in the continuous migration
        model.} From 500 replicate bootstrap datasets, the true (simulated)
        parameter value lies within the distribution of inferred values from
        the bootstrap datasets. Orange line: simulated parameter value. Green
        line: inferred parameter value. Blue histogram: inferred parameter
        values from 500 bootstrap replicate datasets.
    }
    \label{fig:supp-sim-BS-continuous-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-simulated-merger_with_stem_migration-bootstrap-fits.pdf}
    \caption{
        \textbf{Evaluation of confidence intervals in the merger with stem
        migration model.} From 500 replicate bootstrap datasets, the true
        (simulated) parameter value typically lies within the distribution of
        inferred values from the bootstrap datasets. Stem 1 experienced a
        strong bottleneck, at the lower bound of allowed inferred values. In
        this case, and other parameters directly related to Stem 1,
        bootstrap-inferred values do not overlap with the true value, although
        this may be due to incomplete convergence of the optimization at these
        bounds of parameter space. Orange line: simulated parameter value.
        Green line: inferred parameter value. Blue histogram: inferred
        parameter values from 500 bootstrap replicate datasets.
    }
    \label{fig:supp-sim-BS-merger-with-stem-migration}
\end{figure}

\begin{figure}[ht]
    \centering
    \includegraphics{figures/supp-bootstrap-fits-overlap.pdf}
    \caption{
        \textbf{Overlap of bootstrap distributions of inferred parameters
        across multiple parameterized models.} While the details of deep
        history differ substantially between the different inferred models, the
        parameterization of recent history (population sizes, migration rates)
        are similar. Despite the differences in the deep history, many
        parameters related to recent history are consistent across all tested
        models. The single origin model (blue) more often has inferred values
        distinct from other models, although this model provides a far worse
        fit to the data than the other models.
    }
    \label{fig:supp-bootstrap-overlap}
\end{figure}
